Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is an extended abstract of \citet{CompetingBandits-merged}, which in turn subsumes conference papers \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}.}

### 2020-12-20 journal paper/content/sec-bg.tex
Removed some coloring, and a \newpage.
Minor edits re student's comments.

\begin{itemize}
\item \emph{Greedy algorithms} that strive to maximize the reward for the next round given the available information. Thus, they always ``exploit" and never explicitly ``explore".
    \asmargincomment{copied the list from the intro.}

@@

>>>>

\begin{itemize}
\item \emph{Greedy algorithms} that strive to maximize the reward for the next round given the available information. Thus, they always ``exploit" and never explicitly ``explore".
    %\asmargincomment{copied the list from the intro.}

@@

\asedit{Below we discuss which algorithms are better than others, and what does it \emph{mean} for one bandit algorithm to be better than another. This is a rather subtle issue, because some algorithms may be better for some problem instances and/or time intervals, and worse for some others. In particular, ``better" algorithms are better in the long run, but could be worse initially.}

### 2020-12-20 journal paper/content/sec-intro.tex
Minor edits re student's comments.

In this paper, we study the interplay between \exploration and \competition.%
\footnote{\Ie we add \competition to the standard exploration-exploitation tradeoff studied in multi-armed bandits.}
Platforms that engage in exploration typically need to compete against one another. Most importantly, platforms compete for users, who benefit them in two ways:
generate revenue and provide data for learning. This creates a tension:
%between \exploration and \competition.
while exploration may be essential for improving the service tomorrow, it may degrade quality \emph{today}, in which case some of the users can leave and there will be less users to learn from. This may create a ``data feedback loop" when the platform's performance further degrades relative to competitors who keep learning and improving from \emph{their} users, and so forth. Taken to the extreme, such dynamics may cause a ``death spiral" effect when the vast majority of customers eventually switch to competitors.

@@

The main high-level question we ask is:
%that we focus on in this paper is:
%\begin{align}\label{eq:main-Q}
\textbf{How does competition incentivize the adoption of better exploration algorithms?}
%\end{align}
This translates into a number of more concrete questions. While it is commonly assumed that better technology always helps, is this so under competition? Does increased competition lead to higher consumer welfare? How significant are the data feedback loops
%--- when more data leads to more users, which leads to even more data, etc. ---
and how they relate to the anti-trust considerations?
%Finding formalizations that admit meaningful answers is a major part of the overall challenge.
We offer a mix of theoretical results and numerical simulations, in which we study complex interactions between platforms' learning dynamics and users' self-interested behavior.

@@

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a multi-armed bandit algorithm; as such, it can dynamically adjust itself to the observed rewards, but not to the competition. This modeling choice reflects the reality of industrial applications, which build on a huge body of knowledge in machine learning; more on this in Section~\ref{sec:discussion}.
%Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical.
We draw on the machine learning literature to compare bandit algorithms to one another in isolation,%
\footnote{As opposed to ``in competition": as solutions to a stand-alone exploration problem which aims to maximize the total consumer welfare over time (see Appendix~\ref{app:bg} for self-contained background).}
 so that we can talk about \emph{better} bandit algorithms in a principled way.
%\footnote{Such comparisons are somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. More on this in Appendix~\ref{app:bg}. Generally,  ``better" algorithms are better in the long run, but could be worse initially. }
One baseline is algorithms that do not explicitly explore (\emph{greedy algorithms}); in isolation, they are known to perform poorly for a wide variety of problem instances.

@@

\textbf{(iii)} We further soften the decision rule so that the selection probabilities vary smoothly in terms of principals' Bayesian-expected rewards;
%as a function of the difference between  principals' Bayesian-expected rewards;
We call it \SoftMaxRandom, a more realistic middle ground between \HardMax agents and random agents.%
\footnote{Alternatively, one can obtain a \SoftMaxRandom decision rule using a mixture of more ``basic" agent types that follow \HardMax unless the principal's Bayesian-expected rewards are too close to each other.}
In the most technical result of the paper, we find that a better algorithm prevails under much weaker assumptions.
% on what constitutes a better algorithm.
%This is the most technical result of the paper.

>>>>

In this paper, we study the interplay between \exploration and \competition.%
\footnote{\Ie we add \competition to the standard exploration-exploitation tradeoff studied in multi-armed bandits.}
Platforms that engage in exploration typically need to compete against one another. Most importantly, platforms compete for users, who benefit them in two ways:
generating revenue and providing data for learning. This creates a tension:
%between \exploration and \competition.
while exploration may be essential for improving the service tomorrow, it may degrade quality \emph{today}, in which case some of the users can leave and there will be fewer users to learn from. This may create a ``data feedback loop" when the platform's performance further degrades relative to competitors who keep learning and improving from \emph{their} users, and so forth. Taken to the extreme, such dynamics may cause a ``death spiral" effect when the vast majority of customers eventually switch to competitors.

@@

The main high-level question we ask is:
%that we focus on in this paper is:
%\begin{align}\label{eq:main-Q}
\textbf{Whether and how does competition incentivize the adoption of better exploration algorithms?}
%\end{align}
This translates into a number of more concrete questions. While it is commonly assumed that better technology always helps, is this so under competition? Does increased competition lead to higher consumer welfare? How significant are the data feedback loops
%--- when more data leads to more users, which leads to even more data, etc. ---
and how they relate to the anti-trust considerations?
%Finding formalizations that admit meaningful answers is a major part of the overall challenge.
We offer a mix of theoretical results and numerical simulations, in which we study complex interactions between platforms' learning dynamics and users' self-interested behavior.

@@

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a multi-armed bandit algorithm; as such, it can dynamically adjust itself to the observed rewards, but not to the \asedit{competitor's choices, rewards, or market share}.
%\footnote{\gaedit{In particular, the algorithm does not respond to neither the competitors' rewards (which it does not observe) nor its current market share.}} 
This modeling choice reflects the reality of industrial applications, which build on a huge body of knowledge in machine learning; more on this in Section~\ref{sec:discussion}.
%Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical.

\asedit{We draw on the machine learning literature to tell us how to compare bandit algorithms to one another, so that we can talk about \emph{better} bandit algorithms in a principled way. In particular, one considers algorithm's performance \emph{in isolation}: in a standalone exploration problem without competition. We focus on big, qualitative differences in algorithms' performance, see Appendix~\ref{app:bg} for background.}
%We draw on the machine learning literature to compare bandit algorithms to one another in isolation, so that we can talk about \emph{better} bandit algorithms in a principled way.\footnote{\gaedit{This literature typically compares the performance of different algorithms in a stand-alone exploration problem according to their asymptotic regret, which can be interpreted as maximizing consumer welfare in our context (see Appendix~\ref{app:bg} for self-contained background). Thus, we can utilize this comparison measure to assess the quality of the algorithms adopted under competition.}}
%\footnote{Such comparisons are somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. More on this in Appendix~\ref{app:bg}. Generally,  ``better" algorithms are better in the long run, but could be worse initially. }
One baseline is algorithms that do not explicitly explore (\emph{greedy algorithms}); in isolation, they are known to perform poorly for a wide variety of problem instances.

@@

\textbf{(iii)} We further soften the decision rule so that the selection probabilities vary smoothly in terms of principals' Bayesian-expected rewards;
%as a function of the difference between  principals' Bayesian-expected rewards;
We call it \SoftMaxRandom, a more realistic middle ground between \HardMax agents and random agents.%
\footnote{Alternatively, one can obtain a \SoftMaxRandom decision rule using a mixture of more ``basic" agent types that follow \HardMax unless the principal's Bayesian-expected rewards are too close to each other.}
In the most technical result of the paper, we find that a \asedit{sufficiently} better algorithm prevails under much weaker assumptions.
% on what constitutes a better algorithm.
%This is the most technical result of the paper.
